### 2022-10-07 Proposal/proposal.tex
add proposal latex file with basic template

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

%% preamble
\usepackage[margin = 1in]{geometry}

%% metadata
\title{Proposal}
\author{James Horbury\\
    University of Connecticut
}
\date{October 10, 2022}

\begin{document}
\maketitle

\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}
\label{sec:intro}

\section*{Specific Aims}
\addcontentsline{toc}{section}{Specific Aims}
\label{sec:spec}

\section*{Data Description}
\addcontentsline{toc}{section}{Data Description}
\label{sec:data}

\section*{Research Design/Methods/Schedule}
\addcontentsline{toc}{section}{Research Design/Methods/Schedule}
\label{sec:res}

\section*{Discussion}
\addcontentsline{toc}{section}{Discussion}
\label{sec:disc}

\section*{Conclusion}
\addcontentsline{toc}{section}{Conclusion}
\label{sec:conc}

\bibliographystyle{chicago}
\bibliography{citations}

\end{document}

### 2022-10-10 Proposal/proposal.tex
minor edits to proposal
changed title
added content to intro and spec aims

@@
 %% metadata
-\title{Proposal}
+\title{An Analysis of Engine vs. Player Detection in Online Chess using the Jenson-Shannon Divergence}
 \author{James Horbury\\
@@
 
+Recently there has been a very high profile case of alleged cheating in the world of professional chess 
+where the current reigning world chess champion, GM Magnus Carelson, was defeated by GM Hans Niemann, a player 
+that was ranked 42nd in the world and experineced massive growth in FIDE rating in the past two years. This 
+has caused many skeptics in the chess community, including Carelson himself, to suspect cheating may have been at
+play. The ensuing comotion has caused many hobbiest, like myself, to stop and consider how pervasive cheating
+really is in chess, more noteably in online venues, and what methods are employed to flag engine-aided moves.
+
+Current research (link) is being done by Dr. Kenneth R. Regan, of the Computer Science and Engineering 
+Department of the University at Buffalo, employing the use of the Jenson-Shannon Divergence as a measure of 
+similarity between two distributions.
+
 \section*{Specific Aims}
@@
 \label{sec:spec}
+
+Specifically, I would like to investigate how accurate this method of cheat detection is when using a dataset 
+of engine vs. player games as opposed to player vs. player games. This leads me to question the standard at 
+which cheating in chess is measured and whether accusations/bans of alleged cheating are issued justly. I chose 
+to conduct an analysis of this method of cheat detection to promote transparency in this regard.
 

### 2022-10-11 Proposal/proposal.tex
fixed spelling error
finished other sections, added citations

--- 2022-10-07 Proposal/proposal.tex
@@
 \usepackage[margin = 1in]{geometry}
+\usepackage{booktabs}
+\usepackage{natbib}
+\usepackage[colorlinks=true, citecolor=blue]{hyperref}
 
 %% metadata
-\title{Proposal}
+\title{An Analysis of the Use of Centipawn Loss to Detect Cheating in Chess}
 \author{James Horbury\\
@@
 
+Recently there has been a very high profile case of alleged cheating in the world of professional chess 
+where the current reigning world chess champion, GM Magnus Carelson, was defeated by GM Hans Niemann, a player 
+that was ranked 42nd in the world and experineced massive growth in FIDE rating in the past two years. This 
+has caused many skeptics in the chess community, including Carelson himself, to suspect cheating may have been 
+at play. The ensuing comotion has caused many hobbiests to stop and consider how pervasive cheating
+really is in chess, more noteably in online venues, and what methods are employed to flag games with alleged 
+engine-aided moves. This paper will surve as an analysis of the currently available methods of cheat detection 
+and for which cases they are more or less robust.
+
+Current research \citep{regan2011understanding} has been done by Dr. Kenneth R. Regan, of the Computer Science and Engineering 
+Department of the University at Buffalo, employing the use of a model that utilizes the Jenson-Shannon 
+Divergence as a basis for detecting player vs. engine games. However, there is much scrutiny over the 
+transparency of his methods and their effectiveness on different "intensities" of cheating. There are also 
+other open-source options (\href{https://github.com/MGleason1/PGN-Spy}) that can provide an indepth analysis of move-for-move comparisons
+in specific games, an area where Dr. Regan's model appears to be weak.
+
 \section*{Specific Aims}
@@
 \label{sec:spec}
+
+The most widespread methods of cheat detection rely on large amounts of data from the past games of specific 
+players, such as graphing their average centipawn loss per game against their historic FIDE rating. What other 
+alternative methods of cheat detection can be reliably used where the norm falls short?
+
+The purpose of posing this question is to expose the inner-workings of the methodologies behind dectecting 
+cheating in chess in the digital age and provide scientific reasoning to support accusations of cheating in 
+instances where they are valid.
 
@@
 
+I'll be utilizing a dataset of Data for 20,000+ chess games played on Lichess \citep{mysarahmadbhat2021} 
+(including moves, victor, rating, opening details, etc.) to provide insight into how centipawn loss relates 
+to FIDE score in the general population, then do the same using historical data for GM Magnus Carelson 
+\citep{zaidqureshi2022carlsen} and GM Hans Niemann \citep{zaidqureshi2022niemann} to observe any differences. These are user 
+submitted examples through Kaggle, but I should be able to have the same access to the games through the Lichess website.
+
 \section*{Research Design/Methods/Schedule}
@@
 \label{sec:res}
+
+To reitterate, I plan to plot average centipawn loss per game vs. FIDE score for both the general population 
+and for the GM's in question to illustrate the common view that this is a good indicator of cheating in chess. If 
+other examples can be found that follow a growth pattern similar to Niemann's I'll present them also to offer a 
+fair counterargument to the claim that his success is "unnatural". Then, using the opensource move-for-move comparison 
+software mentioned in the Introduction I can go more in depth into how, if possible, Niemann could/could not have 
+cheated in his game against Carlsen. Regarding specific metrics, I'll have to see what the PGN-Spy actually offers in 
+terms of applicable output data and take my analysis from there, which will require more research than what was done 
+for this proposal. Additionally some data cleaning/preparation will be necessary for each Lichess datasets to ensure 
+accuracy.
 
@@
 
+I expect to find that there may be reason to belive that the growth Niemann experience could be considered 
+not outside the bounds of normal given the centipawn loss vs. FIDE score method and that a more in-depth
+analysis must be conducted using other methods of measurement. This would challenge the existing assumptions that
+centipawn loss is a "smoking gun" for cheating and that there are instances it is inaccurate.
+
 \section*{Conclusion}
@@
 \label{sec:conc}
+
+In summary, the goal of this proposal is to analyze the use of comparing centipawn loss to FIDE rating as a metric 
+to detect cheating in chess, and whether move-for-move analysis can support the claim that Niemann likely cheated 
+in his game against Carlsen.
 

### 2022-10-11 Proposal/proposal.tex
fixed hyperlink

@@
 transparency of his methods and their effectiveness on different "intensities" of cheating. There are also 
-other open-source options (\href{https://github.com/MGleason1/PGN-Spy}) that can provide an indepth analysis of move-for-move comparisons
+other \href{https://github.com/MGleason1/PGN-Spy}{open-source options} that can provide an indepth analysis of move-for-move comparisons
 in specific games, an area where Dr. Regan's model appears to be weak.

### 2022-11-07 Manuscript/manuscript.tex
added manuscript sections, edited gitignore

--- 2022-10-10 Proposal/proposal.tex
@@
 \usepackage[margin = 1in]{geometry}
+\usepackage{booktabs}
+\usepackage{natbib}
+\usepackage[colorlinks=true, citecolor=blue]{hyperref}
 
 %% metadata
-\title{An Analysis of Engine vs. Player Detection in Online Chess using the Jenson-Shannon Divergence}
+\title{An Analysis of the Use of Centipawn Loss to Detect Cheating in Chess}
 \author{James Horbury\\
@@
 }
-\date{October 10, 2022}
+\date{November 14, 2022}
 
@@
 \maketitle
+
+\section*{Abstract}
+\addcontentsline{toc}{section}{Abstract}
+\label{sec:abs}
 
@@
 
-Recently there has been a very high profile case of alleged cheating in the world of professional chess 
-where the current reigning world chess champion, GM Magnus Carelson, was defeated by GM Hans Niemann, a player 
-that was ranked 42nd in the world and experineced massive growth in FIDE rating in the past two years. This 
-has caused many skeptics in the chess community, including Carelson himself, to suspect cheating may have been at
-play. The ensuing comotion has caused many hobbiest, like myself, to stop and consider how pervasive cheating
-really is in chess, more noteably in online venues, and what methods are employed to flag engine-aided moves.
-
-Current research (link) is being done by Dr. Kenneth R. Regan, of the Computer Science and Engineering 
-Department of the University at Buffalo, employing the use of the Jenson-Shannon Divergence as a measure of 
-similarity between two distributions.
-
-\section*{Specific Aims}
-\addcontentsline{toc}{section}{Specific Aims}
-\label{sec:spec}
-
-Specifically, I would like to investigate how accurate this method of cheat detection is when using a dataset 
-of engine vs. player games as opposed to player vs. player games. This leads me to question the standard at 
-which cheating in chess is measured and whether accusations/bans of alleged cheating are issued justly. I chose 
-to conduct an analysis of this method of cheat detection to promote transparency in this regard.
-
-\section*{Data Description}
-\addcontentsline{toc}{section}{Data Description}
+\section*{Data}
+\addcontentsline{toc}{section}{Data}
 \label{sec:data}
 
-\section*{Research Design/Methods/Schedule}
-\addcontentsline{toc}{section}{Research Design/Methods/Schedule}
-\label{sec:res}
+\section*{Methods}
+\addcontentsline{toc}{section}{Methods}
+\label{sec:meth}
+
+\section*{Simulation}
+\addcontentsline{toc}{section}{Simulation}
+\label{sec:sim}
+
+\section*{Application}
+\addcontentsline{toc}{section}{Application}
+\label{sec:app}
 
@@
 
-\section*{Conclusion}
-\addcontentsline{toc}{section}{Conclusion}
-\label{sec:conc}
+\section*{Appendix}
+\addcontentsline{toc}{section}{Appendix}
+\label{sec:appx}
 
-\bibliographystyle{chicago}
-\bibliography{citations}
-
-\end{document}
+\section*{References}
+\addcontentsline{toc}{section}{References}
+\label{sec:refs}

### 2022-11-11 Manuscript/manuscript.tex
edited gitignore
added venv to gitignore, updated introduction

@@
 
+% revisit when research is completed
+
 \section*{Introduction}
@@
 
+Recently there has been a very high profile case of alleged cheating in the world of professional chess 
+where the current reigning world chess champion, GM Magnus Carelson, was defeated by GM Hans Niemann, a player 
+that was ranked 42nd in the world and experineced massive growth in FIDE rating in the past two years. This 
+has caused many skeptics in the chess community, including Carelson himself, to suspect cheating may have been 
+at play. The ensuing comotion has caused many hobbiests to stop and consider how pervasive cheating
+really is in chess, more noteably in online venues, and what methods are employed to flag games with alleged 
+engine-aided moves. This paper will surve as an analysis of the currently available methods of cheat detection 
+and for which cases they are more or less robust.
+
+Current research \citep{regan2011understanding} has been done by Dr.
+Kenneth R.
+Regan, of the Computer Science and Engineering Department of the University at Buffalo, employing the use of a model that utilizes the Jenson-Shannon Divergence as a basis for detecting player vs.
+engine games.
+However, there is much scrutiny over the transparency of his methods and their effectiveness on different "intensities" of cheating.
+There are also other \href{https://github.com/MGleason1/PGN-Spy}{open-source options} that can provide an indepth analysis of move-for-move comparisons in specific games, an area where Dr.
+Regan's model appears to be weak.
+
 \section*{Data}
@@
 \label{sec:data}
+
+% insert data from lichess
 
@@
 \label{sec:sim}
+
+% push code from local repo
 
@@
 \label{sec:refs}
+
+\end{document}

### 2022-11-12 Manuscript/manuscript.tex
added more info to introduction

--- 2022-11-07 Manuscript/manuscript.tex
@@
 
+Over the past few months the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen, and "alleged cheating" in chess.
+This "alleged cheating" scandal started when Carlsen withdrew from the Sinquefield Cup, an annual invite-only chess tournament and part of the Grand Chess Tour, after losing to 19-year-old Niemann in a stunning third round defeat.
+
+% Go into detail about why this was such a shock. Include backgrounds of Carlsen and Niemann.
+
+Carlsen didn't immediately stat the reason why he withdrew, in fact at the time he largely stayed quiet.
+Most people correctly assumed though that Carlsen was holding his tongue in calling out Niemann publicly to avoid a potential reatlitory lawsuit for slander.
+It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move.
+
+This has become a matter of significant speculation both inside and outside the chess world given the nature of the situation and the people involved.
+
+The first game was played "over the board" (OTB) in contrast to "virtual" chess which is hosted online.
+Cheating in online chess is very easy.
+One simply needs to run a chess engine (like Stockfish) and execute the oves from their game into the engine as they played, following up with whichever move it suggested you play next.
+Cheating OTB is not as simple.
+Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other elctronic devices with them (see the afore mentioned article about the famous concealed shoe computer).
+
+The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that were made in a single game.
+
 \section*{Data}

### 2022-11-12 Manuscript/manuscript.tex
minor edits to intro

--- 2022-11-07 Manuscript/manuscript.tex
@@
 
+Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen, and “alleged cheating” in chess.
+This “cheating accusation” scandal started when Carlsen withdrew from the Sinquefield Cup after losing to nineteen-year-old Niemann in a stunning third round defeat.
+This was the first time in the then reigning five-time World Chess Champion's career where he withdrew from a major event in progress.
+
+% Go into detail about why this was such a shock. Include backgrounds of Carlsen and Niemann.
+
+Carlsen didn't immediately state the reason as to why he withdrew at first, though many correctly assumed that he was holding his tongue in making any sort of public accusation to avoid a potential retaliatory lawsuit for slander.
+
+It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move, perpetuating rumors of the scandal even further. 
+This has become a matter of significant speculation both inside and outside the chess world given the nature of the situation and the people involved.
+
+The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online.
+Cheating in online chess is very easy.
+One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next.
+Cheating OTB is not as simple.
+Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them (see the afore mentioned article about the famous concealed shoe computer).
+
+The specific allegations related to how Niemann cheated OTB are unknown.
+Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game.
+But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career?
+Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different.
+Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games.
+The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding “easy” vs.
+“difficult” moves.
+
 \section*{Data}

### 2022-11-12 Code/centipawn_metrics_extractor.py
added centipawn metrics code

@@ -1,0 +1,140 @@
+import chess
+import chess.engine
+import chess.pgn
+from datetime import datetime
+import pandas as pd
+import math
+from tqdm import tqdm
+
+def evaluate_game(board, engine, limit):
+   info = engine.analyse(board, limit)
+   return info['score'].white().score(mate_score=1000)
+
+start = datetime.now()
+
+engine = chess.engine.SimpleEngine.popen_uci("C:\Users\james\Python-Dev\stockfish_15_win_x64_avx2\stockfish_15_win_x64_avx2.exe")
+
+movetimesec = 999
+depth = 20
+limit=chess.engine.Limit(time=movetimesec, depth=depth)
+worksheetColumns = ["Date",
+                    "Event Name",
+                    "Event Rounds",
+                    "Round",
+                    "White Name",
+                    "Black Name",
+                    "Result",
+                    "White ELO",
+                    "Black ELO",
+                    "Moves",
+                    "White Av CP Loss",
+                    "Black Av CP Loss",
+                    "Evaluations List",
+                    "White CP Loss List",
+                    "Black CP Loss List",
+                    "PGN",
+                    "Analysis Depth"]
+
+finalDf = pd.DataFrame(columns=worksheetColumns)
+
+f = open("database_file.pgn") # path to your database file in PGN. You can export filtered games from ChessBase, Publish it, download the Html, then download the PNG, and that's it
+
+my_list = []
+
+while True:
+    game = chess.pgn.read_game(f)
+    if game is None:
+        break  # end of file
+
+    my_list.append(game)
+
+
+for game in tqdm(my_list):
+    
+    try:
+        # Evaluate all moves
+        print(game.headers)
+        board = game.board()
+        
+        evaluations = []
+        
+        evaluation = engine.analyse(board, limit=limit)['score'].white().score()
+        evaluations.append(evaluation)
+        
+        for move in game.mainline_moves():
+            board.push(move)
+            positionEvaluation = evaluate_game(board, engine, limit)
+            evaluations.append(positionEvaluation)
+        
+        
+        # Adjust evaluations
+        evaluationsAdjusted = evaluations.copy()
+        evaluationsAdjusted = [max(min(x, 1000), -1000) for x in evaluationsAdjusted]
+        
+        # Calculate metrics
+        white_centipawn_loss_list = []
+        black_centipawn_loss_list = []
[…]

### 2022-11-12 Code/exploratory.py

@@ -1,0 +1,188 @@
+import pandas as pd
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from PIL import Image
+from datetime import datetime
+
+filterByPlayer = True
+
+playerName = "Esipenko"
+playerName = "Firouzja"
+playerName = "Rausis"
+playerName = "Erigaisi"
+playerName = "Gukesh"
+playerName = "Keymer"
+playerName = "Pragg"
+playerName = "Carlsen"
+playerName = "Caruana"
+playerName = "Niemann"
+
+
+filterByMinDate = False
+minDate = "2018-01-01"
+
+filterByMaxDate = False
+maxDate = "2018-01-01"
+
+allPickleFiles = []
+df = pd.DataFrame()
+
+for file in os.listdir():
+    if ".pkl" in file:
+        allPickleFiles.append(file)
+        
+for file in allPickleFiles:
+    localDf = pd.read_pickle(file)
+    concat = [df, localDf]
+    df = pd.concat(concat)
+
+df.drop_duplicates(subset=['Date', 'Event Name', 'Event Rounds', 'Round', 'White Name',
+       'Black Name', 'Result', 'White ELO', 'Black ELO', 'Moves',
+       'White Av CP Loss', 'Black Av CP Loss', 'Analysis Depth'], inplace=True)
+
+
+df = df.dropna(subset=['White Name'], axis=0)
+df = df.dropna(subset=['Black Name'], axis=0)
+df = df.dropna(subset=['White ELO'], axis=0)
+df = df.dropna(subset=['Black ELO'], axis=0)
+df['White ELO'] = df['White ELO'].apply(lambda x: int(x))
+df['Black ELO'] = df['Black ELO'].apply(lambda x: int(x))
+df['Moves'] = df['Moves'].apply(lambda x: int(x))
+df['White Av CP Loss'] = df['White Av CP Loss'].apply(lambda x: float(x))
+df['Black Av CP Loss'] = df['Black Av CP Loss'].apply(lambda x: float(x))
+df['Analysis Depth'] = df['Analysis Depth'].apply(lambda x: int(x))
+df['Date'] = df['Date'].apply(lambda x: str(x).replace(".", "-").replace("??", "01")[:10])
+df['Date'] = df['Date'].apply(lambda x: datetime.strptime(x, "%Y-%m-%d"))
+
+
+# Filter by Player
+if filterByPlayer == True:
[…]

### 2022-11-13 Manuscript/manuscript.tex
added centipawn explanation to intro

@@
 
-% Go into detail about why this was such a shock. Include backgrounds of Carlsen and Niemann.
-
 Carlsen didn't immediately state the reason as to why he withdrew at first, though many correctly assumed that he was holding his tongue in making any sort of public accusation to avoid a potential retaliatory lawsuit for slander.
-
 It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move, perpetuating rumors of the scandal even further.
@@
 Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games.
-The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding “easy” vs.
-“difficult” moves.
+The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs.
+"difficult" moves.
+
+Item 2 requires an assessment of move quality, which is usually done in centipawns.
+The centipawn is a unit of measure used in chess to quantify advantage.
+A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions.
+Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine.
+Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.
 

### 2022-11-14 Manuscript/manuscript.tex
edited intro, added more content to data

@@
 
+Measures of average centipawn loss can be used to evaluate a player's strength and, theoretically, one can compare ACL to one's ELO rating, a measure of one's relative skill level in zero-sum games such as chess, to observe how well they played with respect to this ranking.
+The contribution of this paper is to provide an analysis of this method, as a means to detect cheating in chess, by observing suspicious historic activity (i.e., an individual playing well above their rating in terms of ACL) with the goal of scrutinizing whether it should be treated as a "smoking gun" in terms of evidence of cheating; as was the case with Hans Niemann.
+
+The rest of the paper is organized as follows.
+Section 2 provides specifics about the data used for the ACL versus ELO rating analysis, such as how it was collected and what it does to help answer my research question.
+Section 3 will provide insight into my methods and computer programs used to perform the analysis.
+Section 4 concludes with a discussion.
+
 \section*{Data}
@@
 
-% insert data from lichess
+The data used was provided by Lichess.com, a free and open-source internet chess server that holds the records of many professional games.
+It is a comprehensive list of every professional game Hans Niemann has played, both virtually and OTB, from March of 2018 to September of 2022 and is provided in both .csv and .pkl format.
+Some of the relevant information it contains are what color Niemann and his opponent played and what their respective ELO scores and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.).
+The main information of concern with regard to the entirety of the data is Niemann's ELO score and calculated ACL score of each game of each particular point in time, as this is what will be used to plot.
+All other games that included information where Niemann was not playing were dropped from the original dataset provided by Lichess.
 
-\section*{Methods}
+\section*{Simulation/Methods}
 \addcontentsline{toc}{section}{Methods}
-\label{sec:meth}
-
-\section*{Simulation}
-\addcontentsline{toc}{section}{Simulation}
 \label{sec:sim}

### 2022-11-14 Manuscript/manuscript.tex
minor edits to methods section

@@
 
-% push code from local repo
+% code and resulting figures still in progress, for now just show data provided by ratings.fide.com
+
+After the raw data was cleaned and organized so that only data concerning Niemann's games remained, the CPL of each players moves was calculated using Stockfish, a free and open-source chess engine.
+The sum of these values divided by the total number of moves for each player yielded the ACPL of both Niemann and his opponent for that game.
+No calculation was required for ELO score, since that data was provide in the original dataset.
+Then, in thoery I would have ploted average centipawn loss for each game versus Niemann's then ELO ranking, but the code for this part of my research is still a bit buggy and is not producing the figures that I need to discuss Niemann's sensational jump in ACPL with regard to his current ELO ratings and then compare this to a simlar figure of the standard centipawn loss for every other player in the dataset.
+For now I can show his rating as according to fide.com, the website of the international chess federation that calculates and logs these scores over a players professional history.
 

### 2022-11-14 Manuscript/manuscript.tex
added some figures

@@
 \usepackage[colorlinks=true, citecolor=blue]{hyperref}
+\usepackage{graphicx}
 
 %% metadata
-\title{An Analysis of the Use of Centipawn Loss to Detect Cheating in Chess}
+\title{An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess}
 \author{James Horbury\\
@@
 
+% how to make individual images that are different resolutions the same size?
+
+\includegraphics[scale=0.3]{niemann_rating.png}
+
+\includegraphics[scale=0.5]{carlsen_rating.png}
+
+% report statistical analysis in tables/figures, summarize what they mean
+% discussion to link analysis back to topic
+
 \section*{Discussion}
@@
 \label{sec:disc}
+
+% summarize contributions of research and future directions
+The results of this research largely confirm the claims included in Chess.com's report on Niemann however, although the data does indicate suspicious activity, this is not a "cut and dry" case of cheating, so to speak.
+Observing a single method to detect cheating as we have through average centipawn loss is not enough to confirm for sure that Hans Niemann cheated, both virtually and OTB, more than he claimed but this was never truely the objective of this research.
+The real question posed here was whether historical records of average centipawn loss compared to his ELO ranking could be treated as a "smoking gun" or surefire way to prove cheating alone; which it does not.
+Note: Relevant figures not available, fix this later.
 

### 2022-11-15 Manuscript/manuscript.tex
added citations

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 
@@
 
-% revisit when research is completed
+Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing centipawn math.
+A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science.
+This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become a standard in the chess proctors handbook.
+
+Keywords: Centipawn math, Niemann, Carlsen, Chess
 
@@
 Cheating OTB is not as simple.
-Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them (see the afore mentioned article about the famous concealed shoe computer).
+Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.
 
@@
 Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different.
-Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games.
+Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games \citep{erikallebest2022}.
 The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs.
@@
 % how to make individual images that are different resolutions the same size?
+\begin{figure}
+    \begin{subfigure}{.5\textwidth}
+    \includegraphics[scale=0.2]{niemann_rating.png}
+    \caption{Hans Niemann's ELO ratings over time.}
+    \label{fig:sfig1}
+    \end{subfigure}
 
-\includegraphics[scale=0.3]{niemann_rating.png}
-
-\includegraphics[scale=0.5]{carlsen_rating.png}
+    \begin{subfigure}{.5\textwidth}
+    \includegraphics[scale=0.45]{carlsen_rating.png}
+    \caption{Magnus Carlsen's ELO ratings over time.}
+    \label{fig:sfig2}
+    \end{subfigure}
+\label{fig:fig}
+\end{figure}
 
@@
 % discussion to link analysis back to topic
+
+From the figures provided, one can see that the trends in ELO ranking over time for both Niemann and Carlsen are quite different.
+Although it should be noted that Niemann has had a far shorter professional career than Carlsen, one can observe that he has gone through varying periods of growth and decline whereas Carlsen has been following somewhat of a logarithmic growth pattern.
+Note: compare equal ranges for Carlsen and Niemann as far as ELO ranking for a more unbiased analysis, this may be inaccurate.
 
@@
 
-\section*{Appendix}
-\addcontentsline{toc}{section}{Appendix}
-\label{sec:appx}
-
-\section*{References}
-\addcontentsline{toc}{section}{References}
-\label{sec:refs}
+\bibliographystyle{chicago}
+\bibliography{citations}
 

### 2022-11-15 Manuscript/manuscript.tex
fixed citations

@@
 Cheating in online chess is very easy.
-One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next.
+One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next \citep{kenregan2010}.
 Cheating OTB is not as simple.
@@
 The centipawn is a unit of measure used in chess to quantify advantage.
-A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions.
+A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions .
 Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine.
@@
 
-The data used was provided by Lichess.com, a free and open-source internet chess server that holds the records of many professional games.
+The data used was provided by Lichess.com (via kaggle) \citep{zaidqureshi2022}, a free and open-source internet chess server that holds the records of many professional games.
 It is a comprehensive list of every professional game Hans Niemann has played, both virtually and OTB, from March of 2018 to September of 2022 and is provided in both .csv and .pkl format.
@@
     \includegraphics[scale=0.2]{niemann_rating.png}
-    \caption{Hans Niemann's ELO ratings over time.}
+    \caption{Hans Niemann's ELO ratings over time. \citep{hansniemannrating}}
     \label{fig:sfig1}
@@
     \includegraphics[scale=0.45]{carlsen_rating.png}
-    \caption{Magnus Carlsen's ELO ratings over time.}
+    \caption{Magnus Carlsen's ELO ratings over time. \citep{hansniemannrating}}
     \label{fig:sfig2}

### 2022-12-02 Manuscript/manuscript.tex
fixed title page and abstract page to follow Dr. Yan example
edits to data

@@
-\documentclass[12pt]{article}
+\documentclass[12pt, letterpaper, titlepage]{article}
 \usepackage[utf8]{inputenc}
@@
 \title{An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess}
+
 \author{James Horbury\\
@@
 }
-\date{November 14, 2022}
+\date{\today}
 
@@
 
+\begin{abstract}
 \section*{Abstract}
@@
 
-Keywords: Centipawn math, Niemann, Carlsen, Chess
+\bigskip
+\noindent\sc{Kewords}:
+Centipawn math; 
+Niemann;
+Carlsen;
+Chess;
+\end{abstract}
 
@@
 \label{sec:intro}
+
+How was the topic brought about?
 
@@
 
-The data used was provided by Lichess.com (via kaggle) \citep{zaidqureshi2022}, a free and open-source internet chess server that holds the records of many professional games.
-It is a comprehensive list of every professional game Hans Niemann has played, both virtually and OTB, from March of 2018 to September of 2022 and is provided in both .csv and .pkl format.
-Some of the relevant information it contains are what color Niemann and his opponent played and what their respective ELO scores and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.).
+The dataset was compiled by Rafael Vicente Leite, an independent chess researcher and data scientis, immediately following the events of the scandal with the intent of conducting his own investigation of the Niemann controversy.
+The data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of recorded professional chess games.
+In addition to just holding data from prior games, however, it also provides some engine analysis of these games using Stockfish.
+Liete converted this raw data, that is to say the archive of games, from their initial portable game notation (PGN) format to data that was more amenable to statistical analysis.
+
+Specifically, the processed dataset is comprised of ranked games (both virtual and OTB) of the following professional chess players: Hans Niemann, Erigaisi Arjun, Gukesh D, Alireza Firouzja, Vincent Keymer, and Igors Rausis.
+% I don't know the exact reasoning for why these players were selected aside from Niemann and Carlsen, the two players who sparked the cheating controversy, and Rausis, a known cheater and perhaps "control" of his analysis.
+
+It is a comprehensive list of every professional game they have played from 2018 to 2022 and is provided in both .csv and .pkl format.
+Some of the relevant information it contains are what color was played, what their respective ELO scores were at the time, and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.).
 The main information of concern with regard to the entirety of the data is Niemann's ELO score and calculated ACL score of each game of each particular point in time, as this is what will be used to plot.

### 2022-12-02 Manuscript/manuscript.tex
removed faulty data, and reasonings

@@
 \label{sec:intro}
-
-How was the topic brought about?
 
@@
 
-\section*{Simulation/Methods}
+\section*{Methods}
 \addcontentsline{toc}{section}{Methods}
@@
 
-% code and resulting figures still in progress, for now just show data provided by ratings.fide.com
-
-After the raw data was cleaned and organized so that only data concerning Niemann's games remained, the CPL of each players moves was calculated using Stockfish, a free and open-source chess engine.
-The sum of these values divided by the total number of moves for each player yielded the ACPL of both Niemann and his opponent for that game.
-No calculation was required for ELO score, since that data was provide in the original dataset.
-Then, in thoery I would have ploted average centipawn loss for each game versus Niemann's then ELO ranking, but the code for this part of my research is still a bit buggy and is not producing the figures that I need to discuss Niemann's sensational jump in ACPL with regard to his current ELO ratings and then compare this to a simlar figure of the standard centipawn loss for every other player in the dataset.
-For now I can show his rating as according to fide.com, the website of the international chess federation that calculates and logs these scores over a players professional history.
+% removed faulty methods
 
@@
 
-% how to make individual images that are different resolutions the same size?
-\begin{figure}
-    \begin{subfigure}{.5\textwidth}
-    \includegraphics[scale=0.2]{niemann_rating.png}
-    \caption{Hans Niemann's ELO ratings over time. \citep{hansniemannrating}}
-    \label{fig:sfig1}
-    \end{subfigure}
-
-    \begin{subfigure}{.5\textwidth}
-    \includegraphics[scale=0.45]{carlsen_rating.png}
-    \caption{Magnus Carlsen's ELO ratings over time. \citep{hansniemannrating}}
-    \label{fig:sfig2}
-    \end{subfigure}
-\label{fig:fig}
-\end{figure}
-
-% report statistical analysis in tables/figures, summarize what they mean
-% discussion to link analysis back to topic
-
-From the figures provided, one can see that the trends in ELO ranking over time for both Niemann and Carlsen are quite different.
-Although it should be noted that Niemann has had a far shorter professional career than Carlsen, one can observe that he has gone through varying periods of growth and decline whereas Carlsen has been following somewhat of a logarithmic growth pattern.
-Note: compare equal ranges for Carlsen and Niemann as far as ELO ranking for a more unbiased analysis, this may be inaccurate.
+% removed faulty figures
 

### 2022-12-02 Manuscript/manuscript.tex
refactored dataset

@@
 Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen, and “alleged cheating” in chess.
-This “cheating accusation” scandal started when Carlsen withdrew from the Sinquefield Cup after losing to nineteen-year-old Niemann in a stunning third round defeat.
-This was the first time in the then reigning five-time World Chess Champion's career where he withdrew from a major event in progress.
-
-Carlsen didn't immediately state the reason as to why he withdrew at first, though many correctly assumed that he was holding his tongue in making any sort of public accusation to avoid a potential retaliatory lawsuit for slander.
-It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move, perpetuating rumors of the scandal even further.
-This has become a matter of significant speculation both inside and outside the chess world given the nature of the situation and the people involved.
+The scandal started when Carlsen, the then reining five-time World Chess Champion, withdrew from the Sinquefield Cup after a stunning upset by then nineteen-year-old Niemann in the tournament’s third round.
+Carlsen didn’t immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid any potential retaliatory lawsuit, but it was too late.
+The public was already running wild with the story, and every armature chess player and data scientist was looking for any indication of foul play given the high-stakes nature of the situation and the people involved.
 
@@
 
-% removed faulty methods
+% refactor dataset before adding back methods
 

### 2022-12-04 Manuscript/manuscript.tex
update intro

@@
 
-Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen, and “alleged cheating” in chess.
-The scandal started when Carlsen, the then reining five-time World Chess Champion, withdrew from the Sinquefield Cup after a stunning upset by then nineteen-year-old Niemann in the tournament’s third round.
-Carlsen didn’t immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid any potential retaliatory lawsuit, but it was too late.
-The public was already running wild with the story, and every armature chess player and data scientist was looking for any indication of foul play given the high-stakes nature of the situation and the people involved.
+Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess.
+The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round.
+This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers.
 
-The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online.
-Cheating in online chess is very easy.
-One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next \citep{kenregan2010}.
-Cheating OTB is not as simple.
-Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.
+Carlsen also didn't immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid a retaliatory lawsuit by Niemann while he didn't have proof.
+But it was too late, the public was already running wild with the story and amateur and professional entities alike were launching their own investigations on the matter, looking for any indication of foul play.
 
-The specific allegations related to how Niemann cheated OTB are unknown.
-Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game.
-But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career?
-Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different.
-Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games \citep{erikallebest2022}.
-The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs.
-"difficult" moves.
+Although the question of whether Hans Niemann “cheated” originally stemmed from just the one game, it would be nearly impossible to prove foul play by simply analyzing moves without any form of “smoking gun” in terms of evidence, such as the discovery of a concealed communication device or computer on his person.
+This is primarily due to the modality in which the game was played.
 
-Item 2 requires an assessment of move quality, which is usually done in centipawns.
-The centipawn is a unit of measure used in chess to quantify advantage.
-A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions .
-Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine.
-Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.
+The game between Carlsen and Niemann was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online.
+As one might guess cheating in online chess is quite easy.
+One simply has to rely on outside input, commonly through a chess engine (such as Stockfish, AlphaZero, and MuZero) and execute the moves from their game into the engine as they played, following up with whichever move that was suggested to play next.
+This has become an ever increasing problem as the superiority of top chess engines have been acknowledged ever since IBM's Deep Blue defeated former World Chess Champion Garry Kasparov in 1997.
 
-Measures of average centipawn loss can be used to evaluate a player's strength and, theoretically, one can compare ACL to one's ELO rating, a measure of one's relative skill level in zero-sum games such as chess, to observe how well they played with respect to this ranking.
-The contribution of this paper is to provide an analysis of this method, as a means to detect cheating in chess, by observing suspicious historic activity (i.e., an individual playing well above their rating in terms of ACL) with the goal of scrutinizing whether it should be treated as a "smoking gun" in terms of evidence of cheating; as was the case with Hans Niemann.
+But, cheating OTB is not as simple.
+Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other communication devices with them, making it exceedingly difficult for Niemann to have cheated this way.
+The specific allegations as to how Niemann could have cheated OTB were and currently are still unknown.
+However, since none were found prior, any possibility of this has already been eliminated.
 
-The rest of the paper is organized as follows.
-Section 2 provides specifics about the data used for the ACL versus ELO rating analysis, such as how it was collected and what it does to help answer my research question.
-Section 3 will provide insight into my methods and computer programs used to perform the analysis.
-Section 4 concludes with a discussion.
+One may think the matter would be closed at that, but not quite.
+Maybe we can't divine anything from a single game, but if a trend could somehow be established then, at the very least Carlsen's suspicions could be validated.
+That is to say, if the investigation were to propagate into Niemann's history, where statistical analysis would be possible given the amount of data, one might be able to identify suspicious activity.
+The data is there and Carlsen still holds the public trust, all that needs to be worked out is the how.
+Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
+The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.
+
+Item 2 requires an assessment of move quality which is usually done in centipawns.
+The centipawn is a unit of measure used in chess as a representation of advantage.
+One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss.
+Centipawn loss (CL) is a calculation and numerical score given by a chess engine  (Stockfish, Houdini, Koodo, etc.) as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine.
+In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been.
+A score of 100 means one is losing the equivalent of one pawn per move which would generally be hard to maintain without quickly being mated (unless your opponent misses every chance they get).
 

### 2022-12-04 Code/centipawn_metrics_extractor.py
add updated code and data

@@ -1,140 +1,0 @@
-import chess
-import chess.engine
-import chess.pgn
-from datetime import datetime
-import pandas as pd
-import math
-from tqdm import tqdm
-
-def evaluate_game(board, engine, limit):
-   info = engine.analyse(board, limit)
-   return info['score'].white().score(mate_score=1000)
-
-start = datetime.now()
-
-engine = chess.engine.SimpleEngine.popen_uci("C:\Users\james\Python-Dev\stockfish_15_win_x64_avx2\stockfish_15_win_x64_avx2.exe")
-
-movetimesec = 999
-depth = 20
-limit=chess.engine.Limit(time=movetimesec, depth=depth)
-worksheetColumns = ["Date",
-                    "Event Name",
-                    "Event Rounds",
-                    "Round",
-                    "White Name",
-                    "Black Name",
-                    "Result",
-                    "White ELO",
-                    "Black ELO",
-                    "Moves",
-                    "White Av CP Loss",
-                    "Black Av CP Loss",
-                    "Evaluations List",
-                    "White CP Loss List",
-                    "Black CP Loss List",
-                    "PGN",
-                    "Analysis Depth"]
-
-finalDf = pd.DataFrame(columns=worksheetColumns)
-
-f = open("database_file.pgn") # path to your database file in PGN. You can export filtered games from ChessBase, Publish it, download the Html, then download the PNG, and that's it
-
-my_list = []
-
-while True:
-    game = chess.pgn.read_game(f)
-    if game is None:
-        break  # end of file
-
-    my_list.append(game)
-
-
-for game in tqdm(my_list):
-    
-    try:
-        # Evaluate all moves
-        print(game.headers)
-        board = game.board()
-        
-        evaluations = []
-        
-        evaluation = engine.analyse(board, limit=limit)['score'].white().score()
-        evaluations.append(evaluation)
-        
-        for move in game.mainline_moves():
-            board.push(move)
-            positionEvaluation = evaluate_game(board, engine, limit)
-            evaluations.append(positionEvaluation)
-        
-        
-        # Adjust evaluations
-        evaluationsAdjusted = evaluations.copy()
-        evaluationsAdjusted = [max(min(x, 1000), -1000) for x in evaluationsAdjusted]
-        
-        # Calculate metrics
-        white_centipawn_loss_list = []
-        black_centipawn_loss_list = []
[…]

### 2022-12-04 Code/exploratory.py

@@ -1,188 +1,0 @@
-import pandas as pd
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-from PIL import Image
-from datetime import datetime
-
-filterByPlayer = True
-
-playerName = "Esipenko"
-playerName = "Firouzja"
-playerName = "Rausis"
-playerName = "Erigaisi"
-playerName = "Gukesh"
-playerName = "Keymer"
-playerName = "Pragg"
-playerName = "Carlsen"
-playerName = "Caruana"
-playerName = "Niemann"
-
-
-filterByMinDate = False
-minDate = "2018-01-01"
-
-filterByMaxDate = False
-maxDate = "2018-01-01"
-
-allPickleFiles = []
-df = pd.DataFrame()
-
-for file in os.listdir():
-    if ".pkl" in file:
-        allPickleFiles.append(file)
-        
-for file in allPickleFiles:
-    localDf = pd.read_pickle(file)
-    concat = [df, localDf]
-    df = pd.concat(concat)
-
-df.drop_duplicates(subset=['Date', 'Event Name', 'Event Rounds', 'Round', 'White Name',
-       'Black Name', 'Result', 'White ELO', 'Black ELO', 'Moves',
-       'White Av CP Loss', 'Black Av CP Loss', 'Analysis Depth'], inplace=True)
-
-
-df = df.dropna(subset=['White Name'], axis=0)
-df = df.dropna(subset=['Black Name'], axis=0)
-df = df.dropna(subset=['White ELO'], axis=0)
-df = df.dropna(subset=['Black ELO'], axis=0)
-df['White ELO'] = df['White ELO'].apply(lambda x: int(x))
-df['Black ELO'] = df['Black ELO'].apply(lambda x: int(x))
-df['Moves'] = df['Moves'].apply(lambda x: int(x))
-df['White Av CP Loss'] = df['White Av CP Loss'].apply(lambda x: float(x))
-df['Black Av CP Loss'] = df['Black Av CP Loss'].apply(lambda x: float(x))
-df['Analysis Depth'] = df['Analysis Depth'].apply(lambda x: int(x))
-df['Date'] = df['Date'].apply(lambda x: str(x).replace(".", "-").replace("??", "01")[:10])
-df['Date'] = df['Date'].apply(lambda x: datetime.strptime(x, "%Y-%m-%d"))
-
-
-# Filter by Player
-if filterByPlayer == True:
[…]

### 2022-12-05 Manuscript/manuscript.tex
edited data section

@@
 \usepackage{subcaption}
+\usepackage{setspace}
 
@@
 \maketitle
+
+\doublespace
 
@@
 
-The dataset was compiled by Rafael Vicente Leite, an independent chess researcher and data scientis, immediately following the events of the scandal with the intent of conducting his own investigation of the Niemann controversy.
-The data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of recorded professional chess games.
-In addition to just holding data from prior games, however, it also provides some engine analysis of these games using Stockfish.
-Liete converted this raw data, that is to say the archive of games, from their initial portable game notation (PGN) format to data that was more amenable to statistical analysis.
+Who collected the data?
+The dataset was originally compiled by Rafael Leite, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
+This data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
+Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis.
+He also expanded the data to include the ACPL of each player in each game as computed by Stockfish.
 
-Specifically, the processed dataset is comprised of ranked games (both virtual and OTB) of the following professional chess players: Hans Niemann, Erigaisi Arjun, Gukesh D, Alireza Firouzja, Vincent Keymer, and Igors Rausis.
-% I don't know the exact reasoning for why these players were selected aside from Niemann and Carlsen, the two players who sparked the cheating controversy, and Rausis, a known cheater and perhaps "control" of his analysis.
+What period or range does the data cover?
+The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a grand total of over 10000 games.
+However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
+Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann’s games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to XXXXX)
 
-It is a comprehensive list of every professional game they have played from 2018 to 2022 and is provided in both .csv and .pkl format.
-Some of the relevant information it contains are what color was played, what their respective ELO scores were at the time, and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.).
-The main information of concern with regard to the entirety of the data is Niemann's ELO score and calculated ACL score of each game of each particular point in time, as this is what will be used to plot.
-All other games that included information where Niemann was not playing were dropped from the original dataset provided by Lichess.
+Why does the data help to answer the research question?
+Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
+Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether the hypothesized negative correlation holds true.
 

### 2022-12-09 Manuscript/manuscript.tex
added new edits

@@
 \noindent\sc{Kewords}:
-Centipawn math; 
-Niemann;
-Carlsen;
+Centipawn loss; 
+Hans Niemann;
 Chess;
@@
 
-Who collected the data?
 The dataset was originally compiled by Rafael Leite, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
@@
 
-What period or range does the data cover?
 The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a grand total of over 10000 games.
 However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
-Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann’s games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to XXXXX)
+Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to findreference)
 
-Why does the data help to answer the research question?
 Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/whiteELO-vs-blackELO}
+    \caption{This graph indicates that there is a correlation between Black ELO and White ELO which is due to the structure of the ELO system and tournament modes (i.e.
+swiss) meaning strong players will get paired against strong players}
+    \label{fig:elo_lineplot}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/ELOdistr}
+    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plots.}
+    \label{fig:elo_histplot}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/masterELOdistr}
+    \caption{}
+    \label{fig:elogt23_histplot}
+\end{figure}
+
 \section*{Methods}
 \addcontentsline{toc}{section}{Methods}
-\label{sec:sim}
+\label{sec:meth}
 
-% refactor dataset before adding back methods
+As discussed in the Data section, the original dataset will be filtered so that only values where Hans Niemann's ELO score was greater than or equal to 2300 (i.e., when he qualified as a FIDE Master) will remain.
+This has the effect of weeding out games where Niemann was playing at the junior/amateur level and/or could not qualify for certain tournaments.
+The reason for this was not just to reduce the size and thereby time required for analysis, but also to make sure the data lies strictly within Niemann's professional career.
+This yields a distribution of ELO scores that are more normal than previous.
+
+-	Place plot of new histogram here
+-	To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.
+
+Modeling the relationship between ELO score and ACPL will be done by collapsing over the observations in each discrete bin, calculated previously, to plot an estimate of central tendency accompanied by a confidence interval.
+Assuming linearity, one can theoretically establish a correlation between these two variables and observe whether there was a trend of Niemann playing far better than expected for his then ranking.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/whiteACPL}
+    \caption{}
+    \label{fig:white_acpl}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/blackACPL}
+    \caption{}
+    \label{fig:black_acpl}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/elodiff-higher}
+    \caption{}
+    \label{fig:elodiff-higher}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/elodiff-lower}
+    \caption{}
+    \label{fig:elodiff-lower}
+\end{figure}
 
@@
 
-% removed faulty figures
+To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.
+Collapsing over the observations in each discrete bin to plot an estimate of central tendency along with a confidence interval
+After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result follows expectations.
+That is to say, there appears to be a negative correlation between player ranking and ACPL.
+As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine.
+Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient of about -0.312 for Black and -0.286 for White.
+These r values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent.
+Both values reflect a weak to moderate linear relationship of ELO and ACPL.
+It is possible that a reason to explain why these relationships are not more linear is that when a higher ELO white player get paired against a higher ELO Black player there is greater difficulty for each to find the “strongest” moves available, which may affect their resulting ACPL's.
+To observe this, one can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.
+-	Place plot and observations here
+-	one can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.
+But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, possibly, be instances where he had cheated.
+When Niemann was playing White, the central tendency of his ACPL never appears to move outside the confidence interval of the binned observations in a way that would suggest his playing far exceeded his ranking.
+On the contrary, there seem to be times in which his ACPL was higher than expected such that he was perhaps playing worse than how he “should have”.
+Only when he played Black was there a single instance of binned observations where his ACPL was under the lower bound of that bin's confidence interval.
+Even so, as one can see in the distribution of ELO differences and ACPL of higher and lower rated players, this may not necessarily be due to foul play but instead by simply being paired against a stronger/weaker opponent.
 
@@
 
-% summarize contributions of research and future directions
-The results of this research largely confirm the claims included in Chess.com's report on Niemann however, although the data does indicate suspicious activity, this is not a "cut and dry" case of cheating, so to speak.
-Observing a single method to detect cheating as we have through average centipawn loss is not enough to confirm for sure that Hans Niemann cheated, both virtually and OTB, more than he claimed but this was never truely the objective of this research.
-The real question posed here was whether historical records of average centipawn loss compared to his ELO ranking could be treated as a "smoking gun" or surefire way to prove cheating alone; which it does not.
-Note: Relevant figures not available, fix this later.
+% rewrite findings and improvments
 

### 2022-12-09 Manuscript/manuscript.tex
added revised discussion

@@
 A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science.
-This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become a standard in the chess proctors handbook.
+This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become standard in the chess proctors handbook.
 
@@
 
-The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a grand total of over 10000 games.
-However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
-Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to findreference)
-
-Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
-Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether the hypothesized negative correlation holds true.
-
 \begin{figure}
@@
 
+The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a grand total of over 10000 games.
+However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
+Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to findreference)
+
 \begin{figure}
@@
     \includegraphics[width=\textwidth]{figures/ELOdistr}
-    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plots.}
+    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plot.}
     \label{fig:elo_histplot}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/masterELOdistr}
-    \caption{}
-    \label{fig:elogt23_histplot}
-\end{figure}
+Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
+Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether the hypothesized negative correlation holds true.
 
@@
 
--	Place plot of new histogram here
--	To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/masterELOdistr}
+    \caption{The filtered data uses a more normal dirstribution of ELO scores.}
+    \label{fig:elogt23_histplot}
+\end{figure}
+
+To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Diaconis rule.
+
+\begin{equation}
+    \label{eq:freedman-diaconis}
+    BIN WIDTH = 2 \frac{iqr}{n^{1/3}},
+\end{equation}
+
+where iqr is the interquartile range of the data and n is observations.
 
@@
 Assuming linearity, one can theoretically establish a correlation between these two variables and observe whether there was a trend of Niemann playing far better than expected for his then ranking.
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/whiteACPL}
-    \caption{}
-    \label{fig:white_acpl}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/blackACPL}
-    \caption{}
-    \label{fig:black_acpl}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/elodiff-higher}
-    \caption{}
-    \label{fig:elodiff-higher}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/elodiff-lower}
-    \caption{}
-    \label{fig:elodiff-lower}
-\end{figure}
 
@@
 
-To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Draconis rule.
-Collapsing over the observations in each discrete bin to plot an estimate of central tendency along with a confidence interval
 After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result follows expectations.
 That is to say, there appears to be a negative correlation between player ranking and ACPL.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics{figures/whiteACPL}
+    \caption{Relationship of Niemann's ELO score and ACPL while playing as White.}
+    \label{fig:white_acpl}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics{figures/blackACPL}
+    \caption{Relationship of Niemann's ELO score and ACPL while playing as Black.}
+    \label{fig:black_acpl}
+\end{figure}
+
 As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine.
@@
 To observe this, one can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.
--	Place plot and observations here
--	one can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics{figures/elodiff-higher}
+    \caption{Distribution of ELO difference and ACPL of the higher rated players in the dataset.}
+    \label{fig:elodiff-higher}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics{figures/elodiff-lower}
+    \caption{Distribution of ELO difference and ACPL of the lower rated players in the dataset.}
+    \label{fig:elodiff-lower}
+\end{figure}
+
+One can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.
+
 But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, possibly, be instances where he had cheated.
@@
 
-% rewrite findings and improvments
+The purpose of this research was meant to analyze how one could detect foul play in chess by observing the correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer picture of what many consider to be a common method to identify cheaters, more questions have been raised than answers.
+In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship between his ELO score and ACPL over time.
+Upon further research, it appears that the claims made by many regarding the existence of this relationship is somewhat contested.
+Regrettably, the only real contribution this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether the methodologies used here can even be used to accurately indicate “cheating”.
+The fact that the data used in this paper only reflects the history of a single professional player may be a significant limitation in investigating the true relationship of these variables.
+Shifting the focus from Niemann and performing the same analysis on a dataset consisting of players that were randomly selected from another database may provide more accurate results of what one could expect this relationship could be for the average player.
+Another definite limitation of this study came about as a result during the process of binning observations before plotting the central tendency of ELO score against ACPL.
+It is possible that during this process, the visibility of certain observations that could have been suspicious would undoubtedly be lessened when binned with others that were less so.
+This would have effectively drowned out any times where Niemann could possibly have “cheated” inconsistently.
+Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating.
+Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater.
+A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested.
+This would only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation.
+The possible future directions of this study at the moment would be to investigate this same relationship between ELO score and ACPL for other professional chess players to observe if the same or different effects could be observed.
 

### 2022-12-10 Manuscript/manuscript.tex
changed plot positions
recent edits

@@
 \usepackage{setspace}
+\usepackage{tabularx}
+\usepackage{enumitem}
 
@@
 As one might guess cheating in online chess is quite easy.
-One simply has to rely on outside input, commonly through a chess engine (such as Stockfish, AlphaZero, and MuZero) and execute the moves from their game into the engine as they played, following up with whichever move that was suggested to play next.
-This has become an ever increasing problem as the superiority of top chess engines have been acknowledged ever since IBM's Deep Blue defeated former World Chess Champion Garry Kasparov in 1997.
+One simply has to rely on outside input, commonly through a chess engine (such as Stockfish, Houdini, or Koodo) and execute the moves from their game into the engine as they played, following up with whichever move that was suggested to play next.
+This has become an increasing problem ever since the superiority of top chess engines has been acknowledged with IBM's Deep Blue defeating the former World Chess Champion, Garry Kasparov, in 1997.
 
-But, cheating OTB is not as simple.
-Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other communication devices with them, making it exceedingly difficult for Niemann to have cheated this way.
-The specific allegations as to how Niemann could have cheated OTB were and currently are still unknown.
-However, since none were found prior, any possibility of this has already been eliminated.
+But cheating OTB is not as simple.
+Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other communication devices with them, making it exceedingly difficult for Niemann to have cheated this way but still not impossible.
+The specific allegations as to how Niemann could have cheated OTB were and currently still are unknown.
+However, since no contraband was found prior to the event, any hope of using a “smoking gun” as evidence in the case is futile.
 
 One may think the matter would be closed at that, but not quite.
-Maybe we can't divine anything from a single game, but if a trend could somehow be established then, at the very least Carlsen's suspicions could be validated.
-That is to say, if the investigation were to propagate into Niemann's history, where statistical analysis would be possible given the amount of data, one might be able to identify suspicious activity.
+These days vast swaths of open-access data and high performance computers are increasingly accessible.
+Maybe we can't divine anything from a single game, but if a trend could somehow be established over several hundred or even thousand games then, at the very least, Carlsen's suspicions could be validated.
+That is to say, if the investigation were to propagate into Niemann's history, where statistical analysis would be possible given the amount of data, one might be able to identify suspicious activity if it were done consistently.
 The data is there and Carlsen still holds the public trust, all that needs to be worked out is the how.
+
 Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
@@
 
+Of these cheat detection methods, the particular focus of this paper will be in investigating item 2.
 Item 2 requires an assessment of move quality which is usually done in centipawns.
@@
 One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss.
-Centipawn loss (CL) is a calculation and numerical score given by a chess engine  (Stockfish, Houdini, Koodo, etc.) as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine.
+Centipawn loss (CPL) is a calculation and numerical score given by a chess engine as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine.
+Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine.
 In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been.
-A score of 100 means one is losing the equivalent of one pawn per move which would generally be hard to maintain without quickly being mated (unless your opponent misses every chance they get).
+Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.
 
-\section*{Data}
-\addcontentsline{toc}{section}{Data}
+Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns.
+Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}.
+
+Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths.
+With respect to the Niemann controversy, this hypothesized method has been used by many as an attempt to prove or disprove whether he had “cheated” in the past.
+This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.
+
+The rest of the paper is organized as follows.
+Section~\ref{sec:data} explores the dataset of Hans Niemann's past games that were used for the study.
+Section~\ref{sec:meth} goes over the methods that were used to answer the research question.
+Section~\ref{sec:app} explores the applications of the methods used to investigate the data.
+Finally, Section~\ref{sec:disc} concludes with a discussion and Section~\ref{sec:apx} provides and appendix of technical details.
+
+\section{Data}
 \label{sec:data}
@@
 Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis.
-He also expanded the data to include the ACPL of each player in each game as computed by Stockfish.
+He also expanded the data to include the ACPL of each player in each game as computed by the chess engine, Stockfish.
 
-\begin{figure}
+\begin{figure}[!htb]
     \centering
     \includegraphics[width=\textwidth]{figures/whiteELO-vs-blackELO}
-    \caption{This graph indicates that there is a correlation between Black ELO and White ELO which is due to the structure of the ELO system and tournament modes (i.e.
-swiss) meaning strong players will get paired against strong players}
+    \caption{This graph indicates that there is a correlation between Black ELO and White ELO which is due to the structure of the ELO system, where strong players will get paired against strong players}
     \label{fig:elo_lineplot}
@@
 However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
-Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters (according to findreference)
+Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters.
 
-\begin{figure}
+\begin{figure}[!htb]
     \centering
     \includegraphics[width=\textwidth]{figures/ELOdistr}
-    \caption{From these histogram plots of playing ELO it can be seen that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plot.}
+    \caption{From this histogram it can be observed that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plot.}
     \label{fig:elo_histplot}
@@
 Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
-Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether the hypothesized negative correlation holds true.
+Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether a hypothesized negative correlation holds true within a calculated confidence interval.
 
-\section*{Methods}
-\addcontentsline{toc}{section}{Methods}
+\section{Methods}
 \label{sec:meth}
@@
 
-\begin{figure}
+\begin{figure}[!htb]
     \centering
@@
 
-To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Diaconis rule.
+To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Diaconis rule~\ref{eq:freedman-diaconis}.
 
-\begin{equation}
-    \label{eq:freedman-diaconis}
-    BIN WIDTH = 2 \frac{iqr}{n^{1/3}},
-\end{equation}
-
-where iqr is the interquartile range of the data and n is observations.
+where iqr is the interquartile range of the data and n is the number of observations.
 
@@
 
-\section*{Application}
-\addcontentsline{toc}{section}{Application}
+\section{Application}
 \label{sec:app}
@@
     \centering
-    \includegraphics{figures/whiteACPL}
+    \begin{subfigure}{.5\textwidth}
+      \centering
+      \includegraphics[width=1\linewidth]{figures/whiteACPL}
       \caption{Relationship of Niemann's ELO score and ACPL while playing as White.}
-    \label{fig:white_acpl}
-\end{figure}
-
-\begin{figure}[!htb]
+      \label{fig:sub1}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
       \centering
-    \includegraphics{figures/blackACPL}
+      \includegraphics[width=1\linewidth]{figures/blackACPL}
       \caption{Relationship of Niemann's ELO score and ACPL while playing as Black.}
-    \label{fig:black_acpl}
+      \label{fig:sub2}
+    \end{subfigure}
+    \caption{}
+    \label{fig:test}
 \end{figure}
@@
 As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine.
-Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient of about -0.312 for Black and -0.286 for White.
-These r values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent.
+Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient~\ref{eq:pearson} of about -0.312 for Black and -0.286 for White.
+These \begin{math}r\end{math} values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent.
 Both values reflect a weak to moderate linear relationship of ELO and ACPL.
@@
 
-\begin{figure}[!htb]
+\begin{figure}
     \centering
-    \includegraphics{figures/elodiff-higher}
+    \begin{subfigure}{.5\textwidth}
+      \centering
+      \includegraphics[width=1\linewidth]{figures/elodiff-higher}
       \caption{Distribution of ELO difference and ACPL of the higher rated players in the dataset.}
-    \label{fig:elodiff-higher}
-\end{figure}
-
-\begin{figure}[!htb]
+      \label{fig:higher}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
       \centering
-    \includegraphics{figures/elodiff-lower}
+      \includegraphics[width=1\linewidth]{figures/elodiff-lower}
       \caption{Distribution of ELO difference and ACPL of the lower rated players in the dataset.}
-    \label{fig:elodiff-lower}
+      \label{fig:lower}
+    \end{subfigure}
+    \caption{}
+    \label{fig:elodiff}
 \end{figure}
@@
 
-\section*{Discussion}
-\addcontentsline{toc}{section}{Discussion}
+\section{Discussion}
 \label{sec:disc}
@@
 
+\section{Appendix}
+\label{sec:apx}
+
+Freedman-Diaconis rule defined as follows:
+
+\begin{equation}
+  \label{eq:freedman-diaconis}
+  BIN WIDTH = 2 \frac{iqr}{n^{1/3}},
+\end{equation}
+
+where iqr is the interquartile range of the data and n is the number of observations.
+
+Pearson's equation defined as follows:
+
+\begin{equation}
+  \label{eq:pearson}
+  r = \frac{ \sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y}) }{%
+  \sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2}\sqrt{\sum_{i=1}^{n}(y_i-\bar{y})^2}}
+\end{equation}
+
+where \begin{math}x_{i}\end{math} are the values of the x variable in the sample, \begin{math}\bar{x}\end{math} is the mean of the values of the x variable, \begin{math}y_{i}\end{math} are the values of the y variable in the sample, and \begin{math}\bar{y}\end{math} is the mean of the valus of the y variable.
+
+\begin{table}[h]
+\label{tab:game-theory}
+\begin{tabularx}{\textwidth} { 
+    | >{\centering\arraybackslash}X 
+    | >{\centering\arraybackslash}X 
+    | >{\centering\arraybackslash}X | }
+   \hline
+   Game Theory & Notation & Chess \\
+   \hline
+   a game & G & the game of chess \\
+   \hline
+   a player & \begin{math}i \in \{1, 2\}\end{math} & White or Black \\
+   \hline
+   an action & \begin{math}a_{i} \in A_{i}\end{math} & a move \\
+   \hline
+   a play & \begin{math}\overline{\rm a} \in \overline{\rm A}\end{math} & a single chess game \\
+   \hline
+   a node & \begin{math}x_{j} \in X\end{math} & a position \\
+   \hline
+   a tournament & \begin{math}T(G)\end{math} & a tournament \\
+   \hline
+   AI & \begin{math}v_{i} : X \rightarrow \R\end{math} & a chess engine \\
+   \hline
+   AI best-response & \begin{math}a_{i}^* \in arg max_{a_{i}(x) \in A_{i}(x)}v_{i}(a_{i}(x))\end{math} & best move \\
+   \hline
+\end{tabularx}
+\caption{The terminology of chess as used in game theory.}
+\end{table}
+
 \bibliographystyle{chicago}

### 2022-12-10 Manuscript/manuscript.tex
fixed citations

@@
 The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round.
-This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers.
+This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers\citep{wsj2022}.
 
@@
 
-Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
+Chess.com\citet{chess2022cheating}, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
 The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.
@@
 Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine.
-In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been.
+In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been\citep{avva2022guess}.
 Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.
@@
 Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns.
-Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}.
+Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}\citet{anbarciai}.
 
 Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths.
-With respect to the Niemann controversy, this hypothesized method has been used by many as an attempt to prove or disprove whether he had “cheated” in the past.
+With respect to the Niemann controversy, this hypothesized method has been used by many\citet{medium2022} as an attempt to prove or disprove whether he had “cheated” in the past.
 This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.
@@
 
-The dataset was originally compiled by Rafael Leite, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
-This data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
+The dataset was originally compiled by Rafael Leite\citep{leite2022}, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
+This data\citep{chessbase} itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
 Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis.
@@
 Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating.
-Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater.
+Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater\citep{chessbot}.
 A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested.

### 2022-12-10 Manuscript/manuscript.tex
save current work
minor changes

--- 2022-12-10 Manuscript/manuscript.tex
@@
 The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round.
-This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers.
+This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers \citep{wsj2022}.
 
@@
 
-Chess.com, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
+Chess.com \citep{chess2022cheating}, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
 The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.
@@
 Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine.
-In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been.
+In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been \citep{avva2022guess}.
 Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.
@@
 Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns.
-Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math}.
+Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math} \citep{anbarciai}.
 
 Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths.
-With respect to the Niemann controversy, this hypothesized method has been used by many as an attempt to prove or disprove whether he had “cheated” in the past.
+With respect to the Niemann controversy, this hypothesized method has been used by many \citet{medium2022} as an attempt to prove or disprove whether he had “cheated” in the past.
 This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.
@@
 
-The dataset was originally compiled by Rafael Leite, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
-This data itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
+The dataset was originally compiled by Rafael Leite \citep{leite2022}, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
+This data \citep{chessbase} itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
 Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis.
@@
 Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating.
-Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater.
+Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater \citep{chessbot}.
 A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested.

### 2022-12-10 Manuscript/manuscript.tex
added pagebreak
replaced pagebreak with clearpage
indented text to improve readability

@@
 
-Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing centipawn math.
-A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the sphere of professional chess has launched a conversation about how cheaters can be detected statistically to the forefront of Academic Data Science.
-This paper will explore how one might detect cheating using average centipawn loss and whether these methods are reliable enough to become standard in the chess proctors handbook.
+Evolutions in chess computing have allowed for extensive libraries of past games to be analyzed again utlizing 
+centipawn math. A recent scandal involving the alleged cheating of Chess Grandmaster Hans Niemann in the 
+sphere of professional chess has launched a conversation about how cheaters can be detected statistically 
+to the forefront of Academic Data Science. This paper will explore how one might detect cheating using average
+ centipawn loss and whether these methods are reliable enough to become standard in the chess proctors handbook.
 
@@
 
-Over the past few months, the global chess community has been shaken by a very high-profile story regarding Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess.
-The scandal started when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third round.
-This was the first time in Carlsen's career where he withdrew from a major event in progress, and this fact alone was enough to cause a stir among observers \citep{wsj2022}.
+Over the past few months, the global chess community has been shaken by a very high-profile story regarding 
+Hans Niemann, Magnus Carlsen,  and “alleged cheating” in the world of professional chess. The scandal started
+ when Carlsen, the currently reigning five-time World Chess Champion, unexpectedly withdrew from the 
+ Sinquefield Cup after a stunning upset by the then nineteen-year-old Niemann in the tournament's third 
+ round. This was the first time in Carlsen's career where he withdrew from a major event in progress, and 
+ this fact alone was enough to cause a stir among observers \citep{wsj2022}.
 
-Carlsen also didn't immediately state the reason as to why he withdrew at first, though many correctly assumed he was holding his tongue in making any sort of public accusation of cheating to avoid a retaliatory lawsuit by Niemann while he didn't have proof.
-But it was too late, the public was already running wild with the story and amateur and professional entities alike were launching their own investigations on the matter, looking for any indication of foul play.
+Carlsen also didn't immediately state the reason as to why he withdrew at first, though many correctly 
+assumed he was holding his tongue in making any sort of public accusation of cheating to avoid a retaliatory 
+lawsuit by Niemann while he didn't have proof. But it was too late, the public was already running wild with 
+the story and amateur and professional entities alike were launching their own investigations on the matter, 
+looking for any indication of foul play.
 
-Although the question of whether Hans Niemann “cheated” originally stemmed from just the one game, it would be nearly impossible to prove foul play by simply analyzing moves without any form of “smoking gun” in terms of evidence, such as the discovery of a concealed communication device or computer on his person.
-This is primarily due to the modality in which the game was played.
+Although the question of whether Hans Niemann “cheated” originally stemmed from just the one game, it would be
+ nearly impossible to prove foul play by simply analyzing moves without any form of “smoking gun” in terms of 
+ evidence, such as the discovery of a concealed communication device or computer on his person. This is 
+ primarily due to the modality in which the game was played.
 
-The game between Carlsen and Niemann was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online.
-As one might guess cheating in online chess is quite easy.
-One simply has to rely on outside input, commonly through a chess engine (such as Stockfish, Houdini, or Koodo) and execute the moves from their game into the engine as they played, following up with whichever move that was suggested to play next.
-This has become an increasing problem ever since the superiority of top chess engines has been acknowledged with IBM's Deep Blue defeating the former World Chess Champion, Garry Kasparov, in 1997.
+The game between Carlsen and Niemann was played “over the board” (OTB) in contrast to “virtual chess” which
+ is hosted online. As one might guess cheating in online chess is quite easy. One simply has to rely on 
+ outside input, commonly through a chess engine (such as Stockfish, Houdini, or Koodo) and execute the moves 
+ from their game into the engine as they played, following up with whichever move that was suggested to play
+  next. This has become an increasing problem ever since the superiority of top chess engines has been 
+  acknowledged with IBM's Deep Blue defeating the former World Chess Champion, Garry Kasparov, in 1997.
 
-But cheating OTB is not as simple.
-Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other communication devices with them, making it exceedingly difficult for Niemann to have cheated this way but still not impossible.
-The specific allegations as to how Niemann could have cheated OTB were and currently still are unknown.
-However, since no contraband was found prior to the event, any hope of using a “smoking gun” as evidence in the case is futile.
+But cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are 
+not permitted to bring phones or any other communication devices with them, making it exceedingly difficult
+ for Niemann to have cheated this way but still not impossible. The specific allegations as to how Niemann
+  could have cheated OTB were and currently still are unknown. However, since no contraband was found prior 
+  to the event, any hope of using a “smoking gun” as evidence in the case is futile.
 
-One may think the matter would be closed at that, but not quite.
-These days vast swaths of open-access data and high performance computers are increasingly accessible.
-Maybe we can't divine anything from a single game, but if a trend could somehow be established over several hundred or even thousand games then, at the very least, Carlsen's suspicions could be validated.
-That is to say, if the investigation were to propagate into Niemann's history, where statistical analysis would be possible given the amount of data, one might be able to identify suspicious activity if it were done consistently.
-The data is there and Carlsen still holds the public trust, all that needs to be worked out is the how.
+One may think the matter would be closed at that, but not quite. These days vast swaths of open-access data
+ and high performance computers are increasingly accessible. Maybe we can't divine anything from a single 
+ game, but if a trend could somehow be established over several hundred or even thousand games then, at the
+  very least, Carlsen's suspicions could be validated. That is to say, if the investigation were to propagate 
+  into Niemann's history, where statistical analysis would be possible given the amount of data, one might be 
+  able to identify suspicious activity if it were done consistently. The data is there and Carlsen still holds
+   the public trust, all that needs to be worked out is the how.
 
-Chess.com \citep{chess2022cheating}, which is used by millions of players, has released an expose on their cheat detection methods, specifically with regard to Hans Niemann.
-The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing the time usage for finding “easy” vs “difficult” moves.
+Chess.com \citep{chess2022cheating}, which is used by millions of players, has released an expose on their 
+cheat detection methods, specifically with regard to Hans Niemann. The methods used include (1) comparing 
+moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, 
+(3) comparing player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing 
+the time usage for finding “easy” vs “difficult” moves.
 
-Of these cheat detection methods, the particular focus of this paper will be in investigating item 2.
-Item 2 requires an assessment of move quality which is usually done in centipawns.
-The centipawn is a unit of measure used in chess as a representation of advantage.
-One centipawn is equal to one hundredth of a pawn and, although its value plays no formal role in the game, they are used by chess engines and players alike as a means to evaluate the positions of a chess board via centipawn loss.
-Centipawn loss (CPL) is a calculation and numerical score given by a chess engine as the difference, in centipawns between the move an individual makes and the “strongest” move available at the time as computed by the engine.
-Thus, average centipawn loss (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly speaking, an average assessment of move quality as computed by an engine.
-In practice, ACL is usually somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been \citep{avva2022guess}.
-Although the details of how centipawn loss is calculated will not be explicitly covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory in chess, can be introduced thusly.
+Of these cheat detection methods, the particular focus of this paper will be in investigating item 2. Item 2
+ requires an assessment of move quality which is usually done in centipawns. The centipawn is a unit of measure
+  used in chess as a representation of advantage. One centipawn is equal to one hundredth of a pawn and, 
+  although its value plays no formal role in the game, they are used by chess engines and players alike as a
+   means to evaluate the positions of a chess board via centipawn loss. Centipawn loss (CPL) is a calculation 
+   and numerical score given by a chess engine as the difference, in centipawns between the move an individual 
+   makes and the “strongest” move available at the time as computed by the engine. Thus, average centipawn loss 
+   (ACPL) is simply the average of all the centipawn losses per move over a whole game and is, roughly
+    speaking, an average assessment of move quality as computed by an engine. In practice, ACL is usually 
+    somewhere between 0 and 100, where the closer a player is to 0 the better their moves have been 
+    \citep{avva2022guess}. Although the details of how centipawn loss is calculated will not be explicitly
+     covered in this paper, the basic metric of average centipawn loss, using the terminology of game theory
+      in chess, can be introduced thusly.
 
-Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player \begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and \begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} actual move in centipawns.
-Then the \begin{math}average centipawn loss\end{math} of move \begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math} \citep{anbarciai}.
+Let \begin{math}v_{i}(a_{i}^*(x_{j}))\end{math} be a chess engine's evaluation of the best move for player 
+\begin{math}i\end{math} at position \begin{math}x_{j}\end{math} in centipawns and 
+\begin{math}v_{i}(a_{i}^j(x_{j}))\end{math} be the chess engine's evalution of \begin{math}i's\end{math} 
+actual move in centipawns. Then the \begin{math}average centipawn loss\end{math} of move 
+\begin{math}a_{i}^j(x_{j})\end{math} is defined as \begin{math}v_{i}(a_{i}^*(x_{j})) - v_{i}(a_{i}^j(x_{j}))/n\end{math} 
+where \begin{math}n\end{math} is the total number of moves made by \begin{math}i\end{math} \citep{anbarciai}.
 
-Since the specific details of the methods that Chess.com uses in assessing move quality have not been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified by ELO score, could feasibly serve as the best way to replicate how they would compare a players past performance to their historical strengths.
-With respect to the Niemann controversy, this hypothesized method has been used by many \citet{medium2022} as an attempt to prove or disprove whether he had “cheated” in the past.
-This paper will attempt to investigate whether the use of ACPL as a means to verify the integrity of chess players in the age of modern computing is an appropriate method or just a commonly peddled misconception using data from Hans Niemann's past games.
+Since the specific details of the methods that Chess.com uses in assessing move quality have not 
+been revealed an analysis of how a player's ACPL changes with respect to their ranking, quantified 
+by ELO score, could feasibly serve as the best way to replicate how they would compare a players 
+past performance to their historical strengths. With respect to the Niemann controversy, this hypothesized
+ method has been used by many \citet{medium2022} as an attempt to prove or disprove whether he had “cheated” 
+ in the past. This paper will attempt to investigate whether the use of ACPL as a means to verify the 
+ integrity of chess players in the age of modern computing is an appropriate method or just a commonly 
+ peddled misconception using data from Hans Niemann's past games.
 
-The rest of the paper is organized as follows.
-Section~\ref{sec:data} explores the dataset of Hans Niemann's past games that were used for the study.
-Section~\ref{sec:meth} goes over the methods that were used to answer the research question.
-Section~\ref{sec:app} explores the applications of the methods used to investigate the data.
-Finally, Section~\ref{sec:disc} concludes with a discussion and Section~\ref{sec:apx} provides and appendix of technical details.
+The rest of the paper is organized as follows. Section~\ref{sec:data} explores the dataset of Hans
+ Niemann's past games that were used for the study. Section~\ref{sec:meth} goes over the methods that 
+ were used to answer the research question. Section~\ref{sec:app} explores the applications of the methods 
+ used to investigate the data. Finally, Section~\ref{sec:disc} concludes with a discussion and
+  Section~\ref{sec:apx} provides and appendix of technical details.
 
@@
 
-The dataset was originally compiled by Rafael Leite \citep{leite2022}, a production engineering graduate from the university of Sao Paulo and independent chess researcher, immediately following the events of the initial controversy with the intent of conducting his own investigation.
-This data \citep{chessbase} itself was sourced from ChessBase, a German chess software company that maintains and sells large-scale databases containing the moves of past professional chess games.
-Leite converted this raw data from its original format in portable game notation (PGN) to data that was more amenable to statistical analysis.
-He also expanded the data to include the ACPL of each player in each game as computed by the chess engine, Stockfish.
+The dataset was originally compiled by Rafael Leite \citep{leite2022}, a production engineering graduate from 
+the university of Sao Paulo and independent chess researcher, immediately following the events of the initial 
+controversy with the intent of conducting his own investigation. This data \citep{chessbase} itself was
+ sourced from ChessBase, a German chess software company that maintains and sells large-scale databases 
+ containing the moves of past professional chess games. Leite converted this raw data from its original
+  format in portable game notation (PGN) to data that was more amenable to statistical analysis. He also
+   expanded the data to include the ACPL of each player in each game as computed by the chess engine, Stockfish.
 
@@
     \includegraphics[width=\textwidth]{figures/whiteELO-vs-blackELO}
-    \caption{This graph indicates that there is a correlation between Black ELO and White ELO which is due to the structure of the ELO system, where strong players will get paired against strong players}
+    \caption{This graph indicates that there is a correlation between Black ELO and White ELO which is due to 
+    the structure of the ELO system, where strong players will get paired against strong players}
     \label{fig:elo_lineplot}
@@
 
-The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a grand total of over 10000 games.
-However, for our analysis, we are only interested in the data up to the events of the Sinquefield Cup, which took place on September 4th, 2022.
-Additionally, to avoid having to analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, that is to say the lower bound of most FIDE masters.
+The data covers every official game Hans Niemann has played in his career up to roughly October of 2022, a 
+grand total of over 10000 games. However, for our analysis, we are only interested in the data up to the 
+events of the Sinquefield Cup, which took place on September 4th, 2022. Additionally, to avoid having to 
+analyze an unnecessarily large dataset we will focus on Niemann's games while his ranking was above 2300, 
+that is to say the lower bound of most FIDE masters.
 
@@
     \includegraphics[width=\textwidth]{figures/ELOdistr}
-    \caption{From this histogram it can be observed that the sample for ELOs below 2300 is very low thereby explaining the high fluctuations in the above line plot.}
+    \caption{From this histogram it can be observed that the sample for ELOs below 2300 is very low thereby 
+    explaining the high fluctuations in the above line plot.}
     \label{fig:elo_histplot}
@@
 
-Since we have extensive time series data of the official games Niemann played throughout his career, one can observe the relationship between his rating and ACPL over time.
-Thus, one can observe the correlation coefficient of these two variables assuming linearity, and test whether a hypothesized negative correlation holds true within a calculated confidence interval.
+Since we have extensive time series data of the official games Niemann played throughout his career, one can
+ observe the relationship between his rating and ACPL over time. Thus, one can observe the correlation 
+ coefficient of these two variables assuming linearity, and test whether a hypothesized negative correlation
+  holds true within a calculated confidence interval.
 
@@
 
-As discussed in the Data section, the original dataset will be filtered so that only values where Hans Niemann's ELO score was greater than or equal to 2300 (i.e., when he qualified as a FIDE Master) will remain.
-This has the effect of weeding out games where Niemann was playing at the junior/amateur level and/or could not qualify for certain tournaments.
-The reason for this was not just to reduce the size and thereby time required for analysis, but also to make sure the data lies strictly within Niemann's professional career.
+As discussed in the Data section, the original dataset will be filtered so that only values where Hans 
+Niemann's ELO score was greater than or equal to 2300 (i.e., when he qualified as a FIDE Master) will remain. 
+This has the effect of weeding out games where Niemann was playing at the junior/amateur level and/or could 
+not qualify for certain tournaments. The reason for this was not just to reduce the size and thereby time 
+required for analysis, but also to make sure the data lies strictly within Niemann's professional career. 
 This yields a distribution of ELO scores that are more normal than previous.
@@
 
-To better identify local regions of higher incidence, the number of bins and bin widths have been computed using the Freedman-Diaconis rule~\ref{eq:freedman-diaconis}.
+To better identify local regions of higher incidence, the number of bins and bin widths have been computed
+ using the Freedman-Diaconis rule~\ref{eq:freedman-diaconis}.
 
@@
 
-Modeling the relationship between ELO score and ACPL will be done by collapsing over the observations in each discrete bin, calculated previously, to plot an estimate of central tendency accompanied by a confidence interval.
-Assuming linearity, one can theoretically establish a correlation between these two variables and observe whether there was a trend of Niemann playing far better than expected for his then ranking.
+Modeling the relationship between ELO score and ACPL will be done by collapsing over the observations in 
+each discrete bin, calculated previously, to plot an estimate of central tendency accompanied by a confidence
+ interval. Assuming linearity, one can theoretically establish a correlation between these two variables and
+  observe whether there was a trend of Niemann playing far better than expected for his then ranking.
 
@@
 
-After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result follows expectations.
-That is to say, there appears to be a negative correlation between player ranking and ACPL.
+After plotting the estimated central tendency of ELO score against ACPL, one can observe that the result 
+follows expectations. That is to say, there appears to be a negative correlation between player ranking and 
+ACPL.
 
@@
 
-As discussed in the introduction section at length, it is expected that as the rank of a player increases, they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine.
-Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation coefficient~\ref{eq:pearson} of about -0.312 for Black and -0.286 for White.
-These \begin{math}r\end{math} values need to be calculated separately for each color since their respective ACPL's are closely related within each game, hence they are not independent.
-Both values reflect a weak to moderate linear relationship of ELO and ACPL.
-It is possible that a reason to explain why these relationships are not more linear is that when a higher ELO white player get paired against a higher ELO Black player there is greater difficulty for each to find the “strongest” moves available, which may affect their resulting ACPL's.
-To observe this, one can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.
+As discussed in the introduction section at length, it is expected that as the rank of a player increases, 
+they tend to lose less average centipawns, in effect playing a “stronger” game according to an engine. 
+Calculating the statistics of linearity between ELO and ACPL yields a Pearson correlation 
+coefficient~\ref{eq:pearson} of about -0.312 for Black and -0.286 for White. These \begin{math}r\end{math}
+ values need to be calculated separately for each color since their respective ACPL's are closely related 
+ within each game, hence they are not independent. Both values reflect a weak to moderate linear relationship 
+ of ELO and ACPL. It is possible that a reason to explain why these relationships are not more linear is that 
+ when a higher ELO white player get paired against a higher ELO Black player there is greater difficulty for 
+ each to find the “strongest” moves available, which may affect their resulting ACPL's. To observe this, one 
+ can simply plot the distribution of the ELO difference and ACPL of the higher and lower rated players.
 
@@
 
-One can observe that in the case of the lower rated player the ELO difference matters much more than for the higher ELO player, who seems to have more or less the same accuracy whether playing against a player with similar ELO or a weaker player.
+One can observe that in the case of the lower rated player the ELO difference matters much more than for the
+ higher ELO player, who seems to have more or less the same accuracy whether playing against a player with 
+ similar ELO or a weaker player.
 
-But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, possibly, be instances where he had cheated.
-When Niemann was playing White, the central tendency of his ACPL never appears to move outside the confidence interval of the binned observations in a way that would suggest his playing far exceeded his ranking.
-On the contrary, there seem to be times in which his ACPL was higher than expected such that he was perhaps playing worse than how he “should have”.
-Only when he played Black was there a single instance of binned observations where his ACPL was under the lower bound of that bin's confidence interval.
-Even so, as one can see in the distribution of ELO differences and ACPL of higher and lower rated players, this may not necessarily be due to foul play but instead by simply being paired against a stronger/weaker opponent.
+But the main focus here lies on whether Niemann's performance indicated suspicious behavior which could, 
+possibly, be instances where he had cheated. When Niemann was playing White, the central tendency of his 
+ACPL never appears to move outside the confidence interval of the binned observations in a way that would 
+suggest his playing far exceeded his ranking. On the contrary, there seem to be times in which his ACPL was 
+higher than expected such that he was perhaps playing worse than how he “should have”. Only when he played 
+Black was there a single instance of binned observations where his ACPL was under the lower bound of that
+ bin's confidence interval. Even so, as one can see in the distribution of ELO differences and ACPL of higher 
+ and lower rated players, this may not necessarily be due to foul play but instead by simply being paired 
+ against a stronger/weaker opponent.
 
@@
 
-The purpose of this research was meant to analyze how one could detect foul play in chess by observing the correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer picture of what many consider to be a common method to identify cheaters, more questions have been raised than answers.
-In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship between his ELO score and ACPL over time.
-Upon further research, it appears that the claims made by many regarding the existence of this relationship is somewhat contested.
-Regrettably, the only real contribution this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether the methodologies used here can even be used to accurately indicate “cheating”.
-The fact that the data used in this paper only reflects the history of a single professional player may be a significant limitation in investigating the true relationship of these variables.
-Shifting the focus from Niemann and performing the same analysis on a dataset consisting of players that were randomly selected from another database may provide more accurate results of what one could expect this relationship could be for the average player.
-Another definite limitation of this study came about as a result during the process of binning observations before plotting the central tendency of ELO score against ACPL.
-It is possible that during this process, the visibility of certain observations that could have been suspicious would undoubtedly be lessened when binned with others that were less so.
-This would have effectively drowned out any times where Niemann could possibly have “cheated” inconsistently.
-Which leads to undoubtedly the biggest limitation of this study, possible inability to detect certain styles of cheating.
-Although it might be possible to detect a consistent trend where an individual may be playing far better than expected for their ranking by just observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” cheater \citep{chessbot}.
-A “smart” cheater might prevent themselves from relying on moves suggested by an engine too regularly and may, instead, elect to make only certain moves that were suggested.
-This would only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation.
-The possible future directions of this study at the moment would be to investigate this same relationship between ELO score and ACPL for other professional chess players to observe if the same or different effects could be observed.
+The purpose of this research was meant to analyze how one could detect foul play in chess by observing the 
+correlation between a player's rating and ACPL over a period of time; but rather than painting a clearer 
+picture of what many consider to be a common method to identify cheaters, more questions have been raised than
+ answers. In the case of Hans Niemann, evidently the data does not even suggest a strong linear relationship
+  between his ELO score and ACPL over time. Upon further research, it appears that the claims made by many
+   regarding the existence of this relationship is somewhat contested. Regrettably, the only real contribution 
+   this paper provides is that, in the case of Hans Niemann's professional history, it is inconclusive whether 
+   the methodologies used here can even be used to accurately indicate “cheating”. The fact that the data used 
+   in this paper only reflects the history of a single professional player may be a significant limitation in 
+   investigating the true relationship of these variables. Shifting the focus from Niemann and performing the 
+   same analysis on a dataset consisting of players that were randomly selected from another database may 
+   provide more accurate results of what one could expect this relationship could be for the average player. 
+   Another definite limitation of this study came about as a result during the process of binning observations 
+   before plotting the central tendency of ELO score against ACPL. It is possible that during this process, 
+   the visibility of certain observations that could have been suspicious would undoubtedly be lessened when
+    binned with others that were less so. This would have effectively drowned out any times where Niemann 
+    could possibly have “cheated” inconsistently. Which leads to undoubtedly the biggest limitation of this 
+    study, possible inability to detect certain styles of cheating. Although it might be possible to detect a 
+    consistent trend where an individual may be playing far better than expected for their ranking by just 
+    observing the relationship between ELO score and ACPL, this may not be so for an inconsistent or “smart” 
+    cheater \citep{chessbot}. A “smart” cheater might prevent themselves from relying on moves suggested by 
+    an engine too regularly and may, instead, elect to make only certain moves that were suggested. This would 
+    only give them a slight advantage, but it would be nearly impossible to detect using a simple correlation.
+     The possible future directions of this study at the moment would be to investigate this same relationship
+      between ELO score and ACPL for other professional chess players to observe if the same or different 
+      effects could be observed.
 
@@
 
-where \begin{math}x_{i}\end{math} are the values of the x variable in the sample, \begin{math}\bar{x}\end{math} is the mean of the values of the x variable, \begin{math}y_{i}\end{math} are the values of the y variable in the sample, and \begin{math}\bar{y}\end{math} is the mean of the valus of the y variable.
+where \begin{math}x_{i}\end{math} are the values of the x variable in the sample, \begin{math}\bar{x}\end{math}
+ is the mean of the values of the x variable, \begin{math}y_{i}\end{math} are the values of the y variable in 
+ the sample, and \begin{math}\bar{y}\end{math} is the mean of the valus of the y variable.
 
@@
 
+\clearpage
+
+\pagebreak
 \bibliographystyle{chicago}
